---
title: "Launching the Shiny Apps"
author: "NatureWaves"
date: "2024-11-04"
output: html_document
vignette: >
  %\VignetteIndexEntry{Launching the Shiny Apps}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## How to launch an app
In *Rthoptera*, most of the relevant functions come with a Shiny app companion. To explore which apps are available, run:
```{r}
library(Rthoptera)
launch_app()
```
The `launch_app()` function supports partial matching, which means that, for example, if you want to run the "import_wave" app, you can just type:
```{r}
launch_app("import")
```
and `Rthoptera` will launch the "import_wave" app. However, some apps have longer and similar names, for which you will need to type more characters. For example, if you run:
```{r}
launch_app("temporal_stats")
```
`Rthoptera` will ask you to specify further. If you pay attention to the list of available apps, you will notice that the 'temporal_stats' apps are the only ones ending with 'hq' and 'lq', so if you run:
```{r}
launch_app('hq')
```
*Rthoptera* will understand you are trying to do "launch_app("temporal_stats_hq")". 


